Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2020-06-30 pkg/circtree/inst/wind_paper/draft/JRSSC/rebuttal.tex

We thank you, as well as the associate editor and both reviewers, for the
positive and constructive feedback regarding our manuscript ``\textbf{Circular
Regression Trees and Forests with an Application to Probabilistic Wind
Direction Forecasting}''.

\vspace{5mm}

We have carefully revised our manuscript according to given comments.
The most substantial changes in the manuscript are the following:

\begin{itemize}
\item We have fully revised the `Abstract' in order to put a larger
focus on the application. In addition, we now use the term `application' 
instead of `case study' throughout the manuscript and have also renamed Sect.~4 accordingly.

@@

\item Regarding the employed GLM implemented by Mulder and Klugkist~(2017), we now
explicitly state that the accommodation of circular covariates would be possible
but would require an additional manual adjustment, e.g., an approximation by a finite
Fourier expansion (cf., Lund 1999).

\end{itemize}

\vspace{5mm}

On the following pages a point-to-point response to all comments will be given.
The attached manuscript highlights the changes in the text in blue color.

@@

\vspace{0.5em}
\begin{re}
Page 6, lines 29 and 30: `In general, for additive models a proper model …'?
\end{re}

We have rephrased the sentence according to your suggestion.

\vspace{0.5em}
\begin{re}
Equations (3.1) to (3.4) and elsewhere: I find the use of multiple letters for a single superscript or subscript awkward and prefer to keep lower case letters for indices, so please consider replacing them, for example, T for tree and F for forest. Also for Equation (4.2), how about using parentheses rather than subscripts?
\end{re}

We have followed your suggestions and now use only single letters for super-
and subscripts in the main body of the manuscript. In the appendix, we keep the
notation as it is in order to be consistent with the referenced article by Hothorn~\emph{et al.}~2006.

@@

We first want to thank you for the positive feedback and your valuable input.

We agree with your comment that additional information on the calculation
requirements is valuable for the user. We have therefore listed the needed
computing time of the different methods on a standard desktop computer within
the section 'Computational details'.

\vspace{0.5em}
\begin{re}
Is it possible to include a link to computer code to reproduce your
analysis? This is not essential for publication but would be helpful for
readers who wish to perform similar analyses. The R package you have provided
for your method is very helpful in this regard.  
\end{re}

As we have used data from the Austrian air navigation service provider, which
we are not entitled to distribute, we are unfortunately not able to provide
code and data to reproduce the application. However, as you have already
pointed out the implementation of circular regression trees and forests is
provided in a freely available R-package with examples included in the documentation.
Independent of this publication, we are planning to provide an updated online
platform for the whole `partykit' project in the near future.

@@

\begin{re}
I have very few suggestions to the authors; but here are two:

On page 2, what is the reference for the phrase "Additionally, they
developed a combined ... version ...".
\end{re}

First of all we want to thank you for your positive and constructive feedback.

Thanks for pointing out that the citation here is not clearly assigned. We have therefore
listed the references at the end of the sentence once more.

### 2020-06-30 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Minor corrections by me and Reto
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3022 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e49422c5085c0dbf88805718c78f6726899ff00a

\Abstract{
%SUBMITTED VERSION:
%While circular data occur in a wide range of scientific fields, the methodology
%for distributional modeling and probabilistic forecasting of circular response
%variables is rather limited. Most of the existing methods are built on the
%framework of generalized linear and additive models, which are often
%challenging to optimize and interpret. Therefore, building on previous ideas
%for trees modeling circular means, we suggest a distributional approach for
%regression trees and random forests yielding probabilistic forecasts based on
%the von~Mises distribution. The resulting tree-based models simplify the
%estimation process by using the available covariates for partitioning the data
%into sufficiently homogeneous subgroups so that a simple von~Mises distribution
%without further covariates can be fitted to the circular response in each
%subgroup. These circular regression trees are straightforward to interpret, can
%capture nonlinear effects and interactions, and automatically select the
%relevant covariates that are associated with either location and/or scale
%changes in the von~Mises distribution. Combining an ensemble of circular
%regression trees to a circular regression forest can regularize and smooth the
%covariate effects. The new methods are evaluated in a case study on
%probabilistic wind direction forecasting at two Austrian airports, considering
%other common approaches as a benchmark.
%
%NEW VERSION:
Circular data occur in a wide range of scientific fields, but the methodology
for distributional modeling and probabilistic forecasting of circular response
variables is rather limited. While most of the existing methods are built on
the framework of generalized linear and additive models, which are often
challenging to optimize and interpret, we propose a distributional approach for
regression trees yielding probabilistic forecasts based on the von~Mises
distribution. The resulting tree-based models use the available covariates only
for partitioning the data into sufficiently homogeneous subgroups, allowing for
a simple circular distributional fit in each subgroup. These circular
regression trees are straightforward to interpret, can capture nonlinear
effects and interactions, and automatically select the relevant covariates that
are associated with changes in any distribution parameter. Ensembles or
forests of such circular trees can further improve the forecasts by
regularizing and stabilizing the model. The newly developed methodology is
illustrated in an application on short-term probabilistic wind direction
forecasting for Innsbruck and Vienna Airport in order to direct airplanes to a
safe landing. The predictive skill of the novel approaches is benchmarked with
an additive regression model plus a persistency and a climatology model,
employing the circular continuous ranked probability score.}

@@

Many approaches to model circular data assume that the circular variable of
interest follows a circular distribution, in particular the von~Mises
distribution which is also known as the ``circular normal distribution''. One
of the first regression models with a circular response variable and linear
covariates was presented by \cite{Gould:1969} where the circular mean is
predicted by a linear combination of covariates. \cite{Johnson+Wehrly:1978}
refined this idea by plugging in a link function transforming the linear
predictor to a restricted interval of length~$2\pi$. This generalized linear
model (GLM) type approach was further extended by \cite{Fisher+Lee:1992} and
subsequently by \cite{Fisher:1993} introducing independent GLMs for either the
location or scale parameter with appropriate link functions while keeping the
other parameter constant. Additionally, they developed a combined
heteroscedastic or distributional version with alternating reestimation of the
location and scale parameters conditional on the respective sets of covariates
until convergence \citep{Fisher+Lee:1992, Fisher:1993}. While all of these models are built on well-elaborated
theory, their application in practice remains very challenging, mainly due to
the complexity encountered in optimizing the corresponding log-likelihood
function which is not globally concave. Therefore, highly-informative starting
values are crucial for such circular GLMs to converge
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. In order to avoid
this strong dependence on appropriate initial values,
\cite{Mulder+Klugkist:2017} present a Bayesian alternative of a homoscedastic
GLM for circular data. However, apart from potential difficulties in the
optimization procedure of circular GLMs, the interpretation of the underlying
additive effects is often challenging as well because the link function is
highly nonlinear and the representation of smooth transitions on the unit
circle is not straightforward. For example, the same rotation can be obtained
in either positive or negative direction on the circle leading to an ambiguous
interpretation.

@@

This novel approach to circular regression trees and forests complements the
literature on tree-based circular modeling. \cite{Lund:2002} already introduced a
circular regression tree algorithm where binary splits are made based on an
angular distance measure capturing node homogeneity. However, this only models
changes in the conditional mean, whereas modeling the conditional variance or
full probabilistic distribution would also allow uncertainty in the forecast to
be estimated \citep{Gneiting:2008}. 
%However, this only models
%changes in the conditional mean but not the conditional variance or full
%probabilistic distribution which would allow for also considering uncertainty
%in forecasts \citep{Gneiting:2008}. 
In the field of image based object direction estimation,
\citet{Hara+Chellappa:2017} presented a distribution-based tree-building 
algorithm by applying $k$-means clustering to the target space where
the number of clusters $k$ is adaptively determined by the Bayes
information criterion employing the von~Mises distribution. Despite their
distribution-based splitting approach to circular response variables, they
solely predict the mode of the distribution, thus, providing no full
probabilistic forecasts. Combining these previous ideas for circular trees with the
concept of distributional modeling, we introduce a fully distributional tree
approach that not only considers splits in all distribution parameters, but also yields a
full probabilistic predictive model as proposed by \citet{Gneiting:2008}. In
addition, as a natural extension, ensembles or forests of circular regression
trees are presented in this study, showing that these can further improve
predictive power.

@@

In most generalized linear model (GLM) or generalized additive model (GAM)
approaches to circular regression the location parameter $\mu$ depends on
covariates $\mathbf{z}$ through a link function~$g(\cdot)$, circular intercept
$\mu_0$ and coefficient vector $\boldsymbol{\beta}$: 
%
\begin{equation} \mu = \mu_0 +
g\left(\boldsymbol{\beta}^{\top} \mathbf{z}\right). 
\end{equation} 
%
The link
function transforms the additive predictor to an interval of length $2\pi$.
Typically $g(x) = 2\cdot \text{arctan}(x)$ is employed, as suggested by
\citet{Fisher+Lee:1992}. They also developed a heteroscedastic version by
combining two individual GLMs, each for one of the parameters~$\mu$ and~$\kappa$. 
This provides a first approach to a fully probabilistic regression
model for circular data, albeit the parameters are not regressed simultaneously
on covariates as in the more general framework provided by generalized additive
models for location, scale, and shape \citep[GAMLSS,
][]{Rigby+Stasinopoulos:2005}. Nevertheless, other circular additive models
share the previously discussed difficulties induced by the characteristics of
the log-likelihood function and the strongly nonlinear link function. In general,
for additive models a proper model specification can be very challenging, particularly for a high number of
covariates and no information on possible interactions. Moreover, the additive
structure might impose a smooth effect even if the true underlying effect is an
abrupt shift, which occurs, e.g., in atmospheric wind fields.

@@

    (cf., \citealt{Lund:1999}), we use only the linear components of the lagged

@@

To compare the predictive performance of all proposed models, a circular
analogue of the continuous ranked probability score (CRPS) as introduced by
\cite{Grimit+Gneiting+Berrocal:2006} is computed. Just as the linear version of
the CRPS \citep[for more details see][]{Hersbach:2000} it is a proper scoring
rule \citep{Gneiting+Raftery:2007} and measures the difference between an
observation and the corresponding predicted distribution function in order to
assess the probabilistic goodness of fit for the estimated model. Hence, the
lower the CRPS value the better the predictive performance. Contrary to the
linear version, the circular CRPS reduces not to the absolute error but to the
angular distance when the forecast is deterministic.

@@

%%
\begin{figure}[t]
  \centering
  \setkeys{Gin}{width=1.0\textwidth}
  \includegraphics{_plot_circforest_validation_crpsskill_agg_comparison_with_lowff_v14b.pdf}
  \caption{As Figure~\ref{fig:boxplot_crpsraw}, but showing CRPS skill scores
    with the climatology model as reference. Skill scores are in percent; positive
    values indicate improvements over the reference.} 
  \label{fig:boxplot_crpsskill} 
\end{figure}
%%

@@

To conclude, in general the tree structure can capture nonlinear changes,
shifts, and potential interactions in covariates without prespecification of
such effects. As supported by the application on probabilistic wind direction
forecasting, this can be particularly useful for modeling a highly fluctuating
response, such as typically observed for wind direction, or/and in case of a
large number of possible covariates. Moreover, the application shows that
building ensembles of circular regression trees can even improve the
forecasting performance, as the resulting forests allow for modeling smooth
effects and stabilize the model.

@@

For the circular GLM considered as reference model the corresponding
implementation is provided in the \textsf{R} package \textbf{circglmbayes} by
\cite{Mulder+Klugkist:2017}. In particular the function \code{circGLM} is
applied to estimates the intercept and regression coefficient along with the
concentration parameter.

%\newpage
\bibliography{ref.bib}

>>>>

\Abstract{
%SUBMITTED VERSION:
%While circular data occur in a wide range of scientific fields, the methodology
%for distributional modeling and probabilistic forecasting of circular response
%variables is rather limited. Most of the existing methods are built on the
%framework of generalized linear and additive models, which are often
%challenging to optimize and interpret. Therefore, building on previous ideas
%for trees modeling circular means, we suggest a distributional approach for
%regression trees and random forests yielding probabilistic forecasts based on
%the von~Mises distribution. The resulting tree-based models simplify the
%estimation process by using the available covariates for partitioning the data
%into sufficiently homogeneous subgroups so that a simple von~Mises distribution
%without further covariates can be fitted to the circular response in each
%subgroup. These circular regression trees are straightforward to interpret, can
%capture nonlinear effects and interactions, and automatically select the
%relevant covariates that are associated with either location and/or scale
%changes in the von~Mises distribution. Combining an ensemble of circular
%regression trees to a circular regression forest can regularize and smooth the
%covariate effects. The new methods are evaluated in a case study on
%probabilistic wind direction forecasting at two Austrian airports, considering
%other common approaches as a benchmark.
%
%NEW VERSION:
Circular data occur in a wide range of scientific fields, however the methodology
for distributional modeling and probabilistic forecasting of circular response
variables is rather limited. While most of the existing methods are built within
the framework of generalized linear and additive models, which are often
challenging to optimize and interpret, we propose a distributional approach for
regression trees yielding probabilistic forecasts based on the von~Mises
distribution. The resulting tree-based models use the available covariates 
for partitioning the data into sufficiently homogeneous subgroups, allowing for
a simple circular distributional fit in each subgroup. These circular
regression trees are straightforward to interpret, can capture nonlinear
effects and interactions, and automatically select the relevant covariates that
are associated with changes in any distribution parameter. Ensembles or
forests of such circular trees can further improve the forecasts by
regularizing and stabilizing the model. The newly developed methodology is
illustrated in an application on short-term probabilistic wind direction
forecasting for Innsbruck and Vienna Airport required to direct airplanes to a
safe landing. The predictive skill of the novel approaches is benchmarked with
an additive regression model plus a persistency and a climatological model,
employing the circular continuous ranked probability score.}

@@

Many approaches to model circular data assume that the circular variable of
interest follows a circular distribution, in particular the von~Mises
distribution which is also known as the ``circular normal distribution''. One
of the first regression models with a circular response variable and linear
covariates was presented by \cite{Gould:1969} where the circular mean is
predicted by a linear combination of covariates. \cite{Johnson+Wehrly:1978}
refined this idea by plugging in a link function transforming the linear
predictor to a restricted interval of length~$2\pi$. This generalized linear
model~(GLM) type approach was further extended by \cite{Fisher+Lee:1992} and
subsequently by \cite{Fisher:1993} introducing independent GLMs for either the
location or scale parameter with appropriate link functions while keeping the
other parameter constant. Additionally, they developed a combined
heteroscedastic or distributional version with alternating reestimation of the
location and scale parameters conditional on the respective sets of covariates
until convergence \citep{Fisher+Lee:1992, Fisher:1993}. While all of these models are built on well-elaborated
theory, their application in practice remains very challenging, mainly due to
the complexity encountered in optimizing the corresponding log-likelihood
function which is not globally concave. Therefore, highly-informative starting
values are crucial for such circular GLMs to converge
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. In order to avoid
this strong dependence on appropriate initial values,
\cite{Mulder+Klugkist:2017} present a Bayesian alternative of a homoscedastic
GLM for circular data. However, apart from potential difficulties in the
optimization procedure of circular GLMs, the interpretation of the underlying
additive effects is often challenging as well because the link function is
highly nonlinear and the representation of smooth transitions on the unit
circle is not straightforward. For example, the same rotation can be obtained
in either positive or negative direction on the circle leading to an ambiguous
interpretation.

@@

This novel approach to circular regression trees and forests complements the
literature on tree-based circular modeling. \cite{Lund:2002} already introduced
a circular regression tree algorithm where binary splits are made based on an
angular distance measure capturing node homogeneity. However, this only models
changes in the conditional mean, whereas modeling the conditional variance or
full probabilistic distribution would also allow uncertainty in the forecast to
be estimated \citep{Gneiting:2008}.  In the field of image based object
direction estimation, \citet{Hara+Chellappa:2017} presented a
distribution-based tree-building algorithm by applying $k$-means clustering to
the target space where the number of clusters $k$ is adaptively determined by
the Bayes information criterion employing the von~Mises distribution. Despite
their distribution-based splitting approach to circular response variables,
they solely predict the mode of the distribution, thus, providing no full
probabilistic forecasts. Combining these previous ideas for circular trees with
the concept of distributional modeling, we introduce a fully distributional
tree approach that not only considers splits in all distribution parameters,
but also yields a full probabilistic predictive model as proposed by
\citet{Gneiting:2008}. In addition, as a natural extension, ensembles or
forests of circular regression trees are presented in this study, showing that
these can further improve predictive power.

@@

In most generalized linear model~(GLM) or generalized additive model~(GAM)
approaches to circular regression the location parameter $\mu$ depends on
covariates $\mathbf{z}$ through a link function~$g(\cdot)$, circular intercept
$\mu_0$ and coefficient vector $\boldsymbol{\beta}$: 
%
\begin{equation} \mu = \mu_0 +
g\left(\boldsymbol{\beta}^{\top} \mathbf{z}\right). 
\end{equation} 
%
The link
function transforms the additive predictor to an interval of length $2\pi$.
Typically $g(x) = 2\cdot \text{arctan}(x)$ is employed, as suggested by
\citet{Fisher+Lee:1992}. They also developed a heteroscedastic version by
combining two individual GLMs, each for one of the parameters~$\mu$ and~$\kappa$. 
This provides a first approach to a fully probabilistic regression
model for circular data, albeit the parameters are not regressed simultaneously
on covariates as in the more general framework provided by generalized additive
models for location, scale, and shape \citep[GAMLSS,
][]{Rigby+Stasinopoulos:2005}. Nevertheless, other circular additive models
share the previously discussed difficulties induced by the characteristics of
the log-likelihood function and the strongly nonlinear link function. In general,
for additive models a proper model specification can be very challenging, particularly for a high number of
covariates and no information on possible interactions. Moreover, the additive
structure might impose a smooth effect even if the true underlying effect is an
abrupt shift, which occurs, e.g., in atmospheric wind fields.

@@

    (cf., \citealt{Lund:1999}), we only use the linear components of the lagged

@@

To compare the predictive performance of all proposed models, a circular
analogue of the continuous ranked probability score~(CRPS) as introduced by
\cite{Grimit+Gneiting+Berrocal:2006} is computed. Just as the linear version of
the CRPS \citep[for more details see][]{Hersbach:2000} it is a proper scoring
rule \citep{Gneiting+Raftery:2007} and measures the difference between an
observation and the corresponding predicted distribution function in order to
assess the probabilistic goodness of fit for the estimated model. Hence, the
lower the CRPS value the better the predictive performance. Contrary to the
linear version, the circular CRPS reduces not to the absolute error but to the
angular distance when the forecast is deterministic.

@@

%%
\begin{figure}[t]
  \centering
  \setkeys{Gin}{width=1.0\textwidth}
  \includegraphics{_plot_circforest_validation_crpsskill_agg_comparison_with_lowff_v14b.pdf}
  \caption{As Figure~\ref{fig:boxplot_crpsraw}, but showing CRPS skill scores
    with the climatological model as reference. Skill scores are in percent; positive
    values indicate improvements over the reference.} 
  \label{fig:boxplot_crpsskill} 
\end{figure}
%%

@@

To conclude, in general the tree structure can capture nonlinear changes,
shifts, and potential interactions in covariates without prespecification of
such effects. As supported by the presented application on probabilistic wind direction
forecasting, this can be particularly useful for modeling a highly fluctuating
response, such as typically observed for wind direction, or/and in case of a
large number of possible covariates. Moreover, the application shows that
building ensembles of circular regression trees can even improve the
forecasting performance, as the resulting forests allow for modeling smooth
effects and stabilize the model.

@@

For the circular generalized linear model considered as reference model the corresponding
implementation is provided in the \textsf{R} package \textbf{circglmbayes} by
\cite{Mulder+Klugkist:2017}. In particular the function \code{circGLM} is
applied to estimates the intercept and regression coefficient along with the
concentration parameter.

The computation time on a standard computer is in the order of 10\,milliseconds
for the circular fit (used for the persistency and the climatological model),
seconds for the estimation of the linear model and the circular tree, and from
a few minutes up to an hour for the estimation of the circular forest; the latter
strongly depends on the number of covariates employed and, hence, varies
between the two locations included in the application. However, for a single station and
forecast step, the persistency and linear models must be re-estimated for each
time of interest, and independent climatological models have to be fitted for
all times of interest within a single calendar year, whereas for all times of interest only a single 
circular tree and a single circular forest model is required.

\bibliography{ref.bib}
